---
title: 'DRAFT - fingertipsR: an R package for accessing population health information in England'
author: "Sebastian Fox, Julian Flowers, Simon Thelwall, Daniel Flint, Doris Hain"
output:
  pdf_document: default
  html_notebook: default
  html_document: default
  word_document: default
bibliography: references.bib
---

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(dev="png", 
               dev.args=list(type="cairo"),
               dpi=300)
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
library(fingertipsR)
library(cowplot)
library(Cairo) # for high quality graphics
library(ggplot2)
library(captioner) # for cross-referencing in article
library(stringr)
source("fingertips_paper_captions.R", echo = FALSE)
```

# Abstract

This paper describes the `fingertipsR` package which provides tools accessing a wide range of public health data for England from the [Fingertips website](https://fingertips.phe.org.uk) using its API.

[Fingertips](https://fingertip.phe.org.uk) is a major public repository of population and public health indicators for England, built and maintained by Public Health England. 
The indicators are arranged in thematic or topical profiles covering a wide range of health issues including:

* broad Health Profiles
* specific topics such as liver disease and end of life
* risk factors including alcohol, smoking, physical activity
* population healthcare health services data for general practices, cancer, mental health
* health protection data on general health protection, TB, antimicrobial resistance
* lifecourse profiles for younger and older people
* mortality and morbidity.

Fingertips makes data available for more than 1500 indicators spread across more than 60 profiles. The data can be accessed from <https://fingertips.phe.org.uk> where the data are visualised in variety of ways including heatmaps, choropleth maps, line charts for trends, "spine" charts for comparing multiple indicators for a single area, scatter plots and so on.
Data can be obtained as downloads or figures which can be exported or cut and paste into reports and slides.

A recent addition to the Fingertips platform was an [Automated Programming Interface (API)](https://fingertips.phe.org.uk/api) to enable developers to re-use the data. 
To facilitate access we have designed an *R* package - `fingertipsR` - to facilitate rapid and easy access to the data by analysts and data scientists. 
The package is available from [CRAN](https://cran.r-project.org/).


# Introduction
<!--
 * What fingertips provides
 * What R provides (free open source)
 * Packages as a concept in R
 --> 
 
[Fingertips](https://fingertip.phe.org.uk) is a major public repository of population and public health indicators for England, built and maintained by Public Health England (PHE). 
Data are organised thematically, grouping related indicators together. 
The web interface provides a interactive data visualisations allowing users to examine trends over time, geographical distribution and make comparisons between health providers. 
The web interface also provides a means to download the data behind the visualisations for re-use.
However, accessing the data in this manner limits the user to a single group of indicators for a specified geography per download. 
To enable programmatic access to the data PHE developed an application programming interface (API). 
However, use of an API is highly technical and not always suited to the public health researcher. 

R is free, open source software for statistical analysis. [@rsoftware]
It doubles as both a programming language and analytical environment for performing statistical analyses.
The programming language facilitates expansion of the software through additional user-written 'packages' which are then stored on an online repository. 
Such packages bundle together analytic commands which share a common purpose.
The ease with which R can be expanded has lead to exponential growth in the software, creating a very broad ecosystem of statistical techniques.[citation needed]

The `fingertipsR` package extends R by providing an easy to use access to the Fingertips API allowing direct import of data to R.[@fingertipsrpkg]

# The `fingertipsR` package
<!--
 * Structure of data in fingertips – dates, area types, metadata, value notes, profiles > domains > indicators
 * Accessing the API
 * Description of functions in package
 -->

## The structure of data in fingertips

Public health data gathered and synthesised by PHE are stored on Fingertips in nested thematic groups. 
Profiles group together broad themes of data such as antimicrobial resistance or diabetes. [@johnson2016improving]
These profiles may consist of multiple domains - such as prevalence or targets.
Individual indicators then provide actual values for different measures within the domains for example prevalence among those $\geq$ 65 years of age. 

In addition, indicators can vary by different area types within two broad main geography themes, administrative and health: commissioners of of health care services for local areas (known as Clinical Commissioning Groups or CCGs), hospital groups (acute trusts), upper- and lower-tier local authorities, administrative regions or general practitioner practices among others.
These geographies themselves fall within nested geographies based on whether they are in the administrative or health geography hierarchy. 

All indicators have a fundamentally identical structure; a code that uniquely identifies a geography, three time variables specify the year, quarter and month, variables for count and denominator, where appropriate, provide the count and  denominator, a value column gives the actual value - such as a prevalence or rate - plotted in the various representations on the website. 

## Accessing the fingertips API

Seb/Julian?

## Functions provided by the fingertipsR package

The functions of the fingertipsR package facilitate exploration of the fingertips data in a way that reflects the structure of the data. 
A public health researcher may start by examining which profiles and domains are available
```{r}
library(fingertipsR)
# for common data manipulation functions
library(dplyr, warn.conflicts = FALSE, quietly = TRUE)
ftips_profiles <- profiles(ProfileID = NULL, ProfileName = NULL)
head(ftips_profiles)
```

The researcher could examine what indicators constitute a domain

```{r}
ftips_indicators <- indicators(ProfileID = 8, DomainID = 1000101)
ftips_indicators %>% 
        mutate(IndicatorName = str_trunc(as.character(IndicatorName), width = 20, "right")) %>% 
        head()
```

The researcher may then wish to pull down the data for one or more indicators from one or more domains or profiles.
However, before they can do so, they need to check what geographies are represented by the indicators data are not always available at all geographies.

```{r}
# can't do this at the moment, see issue #21: https://github.com/PublicHealthEngland/fingertipsR/issues/21
```

The researcher is now in a position to read the data into the working environment in R. 

```{r}
ftips_data <- fingertips_data(IndicatorID = 90362, AreaTypeID = 102)
ftips_data %>% 
        select(IndicatorID, IndicatorName, AreaCode, AreaName, AreaType, 
               Sex, Age, Timeperiod, Value, LowerCIlimit, UpperCIlimit) %>%
        mutate(IndicatorName = str_trunc(as.character(IndicatorName), width = 20, "right"),
               AreaName = str_trunc(as.character(AreaName), width = 20, "right")) %>% 
        head()
```

## Using the package to investigate deprivation and life expectancy at birth
 * Perhaps a more involved example of the deprivation/life expectancy analysis used in the vignette

To expand on the functions described above, suppose a researcher wishes to examine the relationship between socio-economic deprivation and life-expectancy at birth.
The fingertipsR package provides a convenient method to extract the data from the fingertips website directly into the working environment of R. 

First, the research loads in the deprivation data at the level of the upper-tier local authorities (also described as County and Unitary Authorities). [@IMD]

```{r}
dep <- deprivation_decile(AreaTypeID = 102, Year = 2015)
head(dep)
```

The researcher can then limit the previously loaded data set giving life expectancy at birth to the level of upper-tier local authority and apply the deprivation data to it.

```{r}
ftips_data <- ftips_data %>%
        # restrict to relevant geography and time
        filter(AreaType == "County & UA" & Timeperiod == "2012 - 14") %>% 
        # merge in deprivation data
        left_join(., dep)
```

The researcher can then plot the relationship between life expectancy at birth and deprivation

<!-- eval=FALSE to print code into paper -->
```{r, eval=FALSE}
p <- ggplot(ftips_data, aes(x = IMDscore, y = Value)) + 
        geom_point() + 
        geom_smooth(se = FALSE, method = "loess") + 
        facet_wrap(~ Sex) + 
        scale_x_reverse("IMD score") + 
        scale_y_continuous("Life expectancy")
p
```


<!-- Then render figure below caption -->
**`r fig_caps("life_exp_birth")`**

```{r, warning=FALSE, echo=FALSE}
p <- ggplot(ftips_data, aes(x = IMDscore, y = Value)) + 
        geom_point() + 
        geom_smooth(se = FALSE, method = "loess") + 
        facet_wrap(~ Sex) + 
        scale_x_reverse("IMD score") + 
        scale_y_continuous("Life expectancy")
p
```

## Extending the package

The Fingertips website, along with its API, are in continuous development. The website has new data added to it once a month, and any changes in the API will be seen then. As a result there is a lot of scope for extension to data available from the R package. 

The package was developed on [GitHub](https://github.com/PublicHealthEngland/fingertipsR), where there is an issues page with public access. This is where bugs and enhancements can be reported. Some extensions to the package that are suggested here are to 

* add a `find_indicator` graphical user interface, to support users that have lesser coding skills
* add a function to identify areas that are significantly worse than England and are significantly trending an the adverse direction
* predict future indicator values based on other indicators within a profile

Users are encouraged to add to the issues list if they feel extensions might be beneficial for the users of this package.

# Discussion/conclusions
* R packages for public health
* Writing R packages as a researcher
* Open source software- github

# References



