---
title: 'fingertipsR: an R package for accessing population health information in England'
author: "Sebastian Fox, Simon Thelwall, Daniel Flint, Doris Hain, Julian Flowers"
output:
  pdf_document: default
  html_notebook: default
  word_document: default
  html_document: default
bibliography: references.bib
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(dev="png", 
               dev.args=list(type="cairo"),
               cache = TRUE,
               dpi=300)
knitr::opts_chunk$set(echo = TRUE, cache = FALSE)
library(fingertipsR)
library(cowplot)
library(Cairo) # for high quality graphics
library(ggplot2)
library(captioner) # for cross-referencing in article
library(stringr)
source("fingertips_paper_captions.R", echo = FALSE)
```

# Abstract

[Fingertips](https://fingertip.phe.org.uk) is a major public repository of population and public health indicators for England, built and maintained by Public Health England (PHE). 
The indicators are arranged in thematic or topical profiles covering a wide range of health issues.
The data can be accessed from <https://fingertips.phe.org.uk> where the data can be visualised using a variety of methods.
Data can also be exported for a group of indicators, for a single geography per download. 

A recent addition to the Fingertips platform was an [Application Programming Interface](https://fingertips.phe.org.uk/api) (API) to enable developers to re-use the data. 
To facilitate access to Fingertips data we have developed an R package - `fingertipsR` - enabling rapid and easy access to fingertips data by analysts and data scientists. 
The package is available from the [Comprehensive R Archive Network](https://cran.r-project.org/) (CRAN).

This paper describes the Fingertips API and the use of the `fingertipsR` package.

# Introduction
<!--
 * What fingertips provides
 * What R provides (free open source)
 * Packages as a concept in R
 --> 
 
[Fingertips](https://fingertip.phe.org.uk) is a major public repository of population and public health indicators for England, built and maintained by Public Health England (PHE). 
Data are organised thematically, grouping related indicators together. 
The web interface provides an interactive platform for data visualisations allowing users to examine trends over time, geographical distribution and make comparisons between healthcare providers. 
The web interface also provides a means to download the data behind the visualisations for re-use.
However, accessing the data in this manner limits the user to a single group of indicators within for a specified geography per download, creating a time burden on the researchers. 
To enable programmatic access to the data PHE developed an application programming interface (API). 
In contrast to the web interface, the API allows a researcher to extract data for many different indicators across domains and for multiple geographies. 
However, use of an API is highly technical and not always suited to the public health researcher. 

R is a free, open source software for statistical analysis. [@rsoftware]
It doubles as both a programming language and analytical environment for performing statistical analyses.
The programming language facilitates expansion of the software through additional user-written 'packages' which are then stored in an online repository. 
Such packages bundle together analytic commands which share a common purpose.
The ease with which R can be expanded through its packaging system has led to exponential growth in the software, creating a very broad ecosystem of statistical techniques. [@johnfox2009]

The `fingertipsR` package extends R by providing an easy-to-use set of functions to query the Fingertips API, allowing direct import of data to R. [@fingertipsrpkg]

# The `fingertipsR` package
<!--
 * Structure of data in fingertips – dates, area types, metadata, value notes, profiles > domains > indicators
 * Accessing the API
 * Description of functions in package
 -->

## The structure of data in Fingertips

Public health data gathered and synthesised by PHE are stored on Fingertips in nested thematic groups. 
Profiles group together broad themes of data such as antimicrobial resistance or diabetes. [@johnson2016improving]
These profiles may consist of multiple domains bundling together common indicators within a theme - such as prevalence of disease or primary care treatment targets. <!-- e.g. diabetes profile -->
Individual indicators then provide actual values for different measures within the domains, for example prevalence of dementia among those $\geq$ 65 years of age. 
In addition, indicators can vary by different area types within two broad geographic themes, administrative and health.
Health geographies include commissioners of health care services for local areas (known as Clinical Commissioning Groups or CCGs), hospital groups (acute trusts) and general practices. 
Administrative geographies include upper- and lower-tier local authorities and government regions among others.
These geographies themselves fall within nested geographies according to whether they are in the administrative or health geography hierarchy. 

All indicators have a fundamentally identical structure. 
Columns provide: a code that uniquely identifies a geography, four time variables that specify year, quarter, month and year range, counts and denominators (data are provided where appropriate) and a value column gives the actual value - such as a prevalence or rate - to be plotted in the various representations on the website. 
Rows provide an observation for the indicator at a specified time period, age group and sex for a given geography.

## Accessing the Fingertips API

The API aims to provide the same data as can be visualised on the Fingertips web site through a collection of RESTful web services. 
The API was created according to open data principles so that the data on the Fingertips site could be available for anyone to access, use or share. [@ahmed] 
The web-based API enables the data to be accessed from any location over the internet using the researcher's programming language of choice without any knowledge required of the system that provides that data. 
The data are transferred in JavaScript Object Notation (JSON), a lightweight data format that is intended to be easy for both humans to read and machines to generate and parse.

## Functions provided by the fingertipsR package

The functions of the fingertipsR package facilitate exploration of the fingertips data in a way that reflects the structure of the data. 
A public health researcher may start by examining which profiles and domains are available:
```{r}
# load the fingertipsR package
library(fingertipsR)

# load dplyr for common data manipulation functions
library(dplyr, warn.conflicts = FALSE, quietly = TRUE)

ftips_profiles <- profiles(ProfileID = NULL, ProfileName = NULL)
head(ftips_profiles)
```

The researcher could examine what indicators constitute a domain

```{r}
ftips_indicators <- indicators(DomainID = 1000101)
ftips_indicators %>% 
        head
```

The researcher may then wish to retrieve data for one or more indicators from one or more domains or profiles.
However, before they can do so, they need to check what geographies are represented as the indicators data are not always available for all geographies.

```{r}
ind_at <- indicator_areatypes() %>%
        left_join(area_types(), by = c("AreaTypeID","AreaTypeID"))  %>%
        select(IndicatorID, AreaTypeID, AreaTypeName) %>%
        unique()
head(ind_at)
```

The researcher is now in a position to read the data into the working environment in R. 

```{r}
ftips_data <- fingertips_data(IndicatorID = 90366, AreaTypeID = 102)
ftips_data %>% 
        select(IndicatorID, IndicatorName, AreaCode, AreaName, AreaType, 
               Sex, Age, Timeperiod, Value, LowerCI95.0limit, UpperCI95.0limit) %>%
        head()
```

## Using the package to investigate deprivation and life expectancy at birth

To expand on the functions described above, suppose a researcher wishes to examine the relationship between socio-economic deprivation and life-expectancy at birth.
The `fingertipsR` package provides a convenient method to extract the data from the fingertips website directly into the working environment of R.

The Index of Multiple Deprivation is a score based on 7 domains of relative deprivation. 
These domains are: Income Deprivation, Employment Deprivation, Education, Skills and Training Deprivation, Health Deprivation and Disability, Crime, Barriers to Housing and Services, Living Environment Deprivation. [@IMD]

First, the researcher loads in the deprivation data for the upper-tier local authorities geographies (also described as County and Unitary Authorities).

```{r}
dep <- deprivation_decile(AreaTypeID = 102, Year = 2015)
head(dep)
```

The researcher can then limit the previously loaded data set giving life expectancy at birth to the level of upper-tier local authority and merge the deprivation data onto it.

```{r, message=FALSE, warning=FALSE}
ftips_data <- ftips_data %>%
        # restrict to relevant geography and time
        filter(AreaType == "County & UA" & Timeperiod == "2012 - 14") %>% 
        # merge in deprivation data
        left_join(., dep)
```

The researcher can then plot the relationship between life expectancy at birth and deprivation.

<!-- eval=FALSE to print code into paper -->
```{r, eval=FALSE}
p <- ggplot(ftips_data, aes(x = IMDscore, y = Value)) + 
        geom_point() + 
        geom_smooth(se = FALSE, method = "loess") + 
        facet_wrap(~ Sex) + 
        scale_x_reverse("IMD score") + 
        scale_y_continuous("Life expectancy at birth (years)")
p
```


<!-- Then render figure below caption -->
**`r fig_caps("life_exp_birth")`**

```{r, warning=FALSE, echo=FALSE}
p <- ggplot(ftips_data, aes(x = IMDscore, y = Value)) + 
        geom_point() + 
        geom_smooth(se = FALSE, method = "loess") + 
        facet_wrap(~ Sex) + 
        scale_x_reverse("IMD score") + 
        scale_y_continuous("Life expectancy")
p
```

## Extending the package

The Fingertips website and API are under continuous development and improvement. 
The website has new data added to it on a monthly schedule, and any changes to the API are implemented at the same time. 
As a result there is a lot of scope for extension to the functions that could be provided in the R package. 

The package was developed openly on [GitHub](https://GitHub.com/PublicHealthEngland/fingertipsR), an online archive where computer code is stored, and in which it is carefully version controlled through Git software.
Various features of Git and GitHub facilitate scientific collaboration. [@ram2013git, @Bryan2017] 

GitHub is a decentralised system - copies of code can be stored locally (i.e. on a computer hard drive) and remotely (as *forks* on the GitHub website) then synchronised at the main repository.
Crucially, Git and GitHub allow any user to examine and adapt code then provide the adapted code back to the main repository for a project.
Additionally, GitHub provides an issue tracker with which users of software can report problems with the software or suggest new features that might be incorporated. 
The version control features identify all changes that were made to analytic code, who made them and at what time. 
This allows analytic code to be reverted to previous states, meaning that breaking changes can be easily reverted. 
Feedback based on the original release of the package has led to developments to support different types of researchers, including: 

* add a `select_indicators()` graphical user interface function, to support users less comfortable with R's square bracket indexing
* add a function to identify areas that are significantly worse than the national average and show a significant trend towards worse values

Additional suggestions for future updates include:

 * predict future indicator values based on other indicators within a profile
 * detect extreme values 
 * flag new or recently updated indicators

Users of the `fingertipsR` package are encouraged to add to the issues list if they feel extensions might be beneficial for the users of this package.
At the date of writing, the package itself has been downloaded from CRAN 607 times, 30 issues have been raised at GitHub and four researchers have contributed to the project repository. 

# Discussion/conclusions

The work presented here illustrates a number of new and important concepts in public health research. 
Software for public health can be developed by researchers who are not professional software developers and can be distributed freely and easily on the web. 
Open source software (software where the code can be examined and adapted) allows the rapid development of new functionality, greatly expanding the uses to which it can be put and rapidly providing solutions to otherwise unmet need. 
The decentralised nature of the version control software Git means that scientific software and analytic code can be developed without the need for physical co-location of collaborators - some of the authors of this paper have never physically met. 
Open data facilitates the sharing of information important to public health. We have found that there is a strong need and demand for quality data to be openly available for public health professionals and researchers who may use a variety technologies to access it. Making the data available in technology-neutral and standard formats reduces barriers to its reuse and increases the value of the data as it can then be employed in novel contexts. 
Together, open data, open source software and distributed software development have a great deal to offer public health. 

# References



