---
title: "Plotting healthy life expectancy and life expectancy by deprivation for English local authorities"
author: "Seb Fox"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Life expectancy by deprivation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This worked example attempts to document a common workflow a user might follow when using the `fingertips` package.

Suppose you want to plot healthy life expectancy and life expectancy by deprivation for a given year of data that fingertips contains - you will begin by wondering *where to start*.

## Where to start

There is one function in the `fingertips` package that extracts data from the Fingertips API: `fingertips_data()`. This function has the following inputs:

* IndicatorID
* AreaCode
* DomainID
* ProfileID
* AreaTypeID (this defaults to 102; County and Unitary Authority)
* ParentAreaTypeID (this defaults to 6; Government Office Region)

One of *IndicatorID*, *DomainID* or *ProfileID* must be complete. *AreaCode* needs completion if you are extracting data for a particular area or group of areas only. *AreaTypeID* determines the geography to extract the data for. In this case we want County and Unitary Authority level. *ParentAreaTypeID* requires an area type code that the *AreaTypeID* maps to.

Therefore, the inputs to the `fingertips_data` function that we need to find out are the ID codes for:

* IndicatorID 
* AreaTypeID
* ParentAreaTypeID

We need to begin by calling the `fingertips` package: 
```{r packages, echo=FALSE}
library(fingertips)
library(ggplot2)
```

## IndicatorID

There are two indicators we are interested in for this exercise. Without consulting the [Fingertips website](https://fingertips.phe.org.uk/  "Fingertips"), we know approximately what they are called:

* Healthy life expectancy
* Life expectancy

We can use the `indicators()` function to return a list of all the indicators within Fingertips. We can then filter the name field for the term *life expectancy* (note, the IndicatorName field has been converted to lower case in the following code chunk to ensure matches will not be overlooked as a result of upper case letters).

```{r indicators, cache=TRUE}
inds <- indicators()
life_expectancy <- inds[grepl("life expectancy", tolower(inds$IndicatorName)),]

# Because the same indicators are used in multiple profiles, there are many repeated indicators in this table (some with varying IndicatorName but same IndicatorID)

# This returns a record for each IndicatorID
life_expectancy <- unique(life_expectancy[duplicated(life_expectancy$IndicatorID) == FALSE,
                                          c("IndicatorID", "IndicatorName")]) 
knitr::kable(life_expectancy, row.names = FALSE)
```

The two indicators we are interested in from this table are:

* 90362
* 90366

## AreaTypeID

We can work out what the *AreaTypeID* codes we are interested in using the function `area_types()`. We've decided that we want to produce the graph at County and Unitary Authority level. From the section [Where to start] we need codes for *AreaTypeID* and *ParentAreaTypeID.*

```{r area type}
areaTypes <- area_types()
DT::datatable(areaTypes, filter = "top", rownames = FALSE)

```

The table shows that the *AreaID* for County and Unitary Authority level is 102. The third column, *ParentAreaID*, shows the IDs of the area types that these map to. In the case of County and Unitary Authorities, these are:

```{r dist, echo=FALSE}
knitr::kable(areaTypes[areaTypes$AreaID == 102, c("ParentAreaID","ParentAreaName")], row.names = FALSE)
```

*ParentAreaID* is 6 by default for the `fingertips_data()` function, so we can stick with that.

## Deprivation

We want to plot life expectancy against deprivation information. The package has a `deprivation_deciles()` function that allows us to return this information. This is populated from the CIPFA Indices of Multiple Deprivation (IMD). Note, there is only information for upper and lower tier local authorities (*AreaTypeID* = 102 and 101 respectively). IMD has only been produced for the years 2010 and 2015.

```{r deprivation}
dep <- deprivation_decile(AreaTypeID = 102, Year = 2015)
DT::datatable(dep, filter = "top", rownames = FALSE)
```

## Extracting the data

Finally, we can use the `fingertips_data()` function with the inputs we have determined previously.

```{r extract}
indicators <- c(90362, 90366)
data <- fingertips_data(IndicatorID = indicators,
                        AreaTypeID = 102)

pander::pandoc.table(tail(data), 
                     style="rmarkdown",
                     split.tables = 90, 
                     keep.line.breaks = TRUE)
```

The data frame returned by `fingertips_data()` contains 21 variables.  For this exercise, we are only interested in a few of them:

* IndicatorID
* AreaCode
* Sex
* Timeperiod
* Value

```{r refine variables}
cols <- c("IndicatorID", "AreaCode", "Sex", "Timeperiod", "Value")
data <- data[data$AreaType == "County & UA" & data$Timeperiod == "2012 - 14", cols]

# merge deprivation onto data
data <- merge(data, dep, by.x = "AreaCode", by.y = "AreaCode", all.x = TRUE)

# remove NA values
data <- data[complete.cases(data),]
DT::datatable(data, filter = "top", rownames = FALSE)
```

## Plotting outputs

Using `ggplot2` it is possible to plot the outputs

```{r plot, fig.width=8, fig.height=5}
p <- ggplot(data, aes(x = IMDscore, y = Value, col = factor(IndicatorID)))
p <- p + 
        geom_point() +
        geom_smooth(se = FALSE, method = "loess") +
        facet_wrap(~ Sex) +
        scale_colour_manual(name = "Indicator",
                            breaks = c("90366", "90362"),
                            labels = c("Life expectancy", "Healthy life expectancy"),
                            values = c("#128c4a", "#88c857")) +
        scale_x_reverse() + 
        labs(x = "IMD deprivation",
             y = "Age",
             title = "Life expectancy and healthy life expectancy at birth \nfor Upper Tier Local Authorities (2012 - 2014)") +
        theme_bw()
print(p)
```


